Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2021-11-22 paper/paper.tex
Brings the description of the model into the chemoaffinity section of the results

% Ability to use colour in text
\usepackage[usenames]{color}

@@

\textbf{New introduction:} Simpson and Goodhill presented a model in which the
chemoaffinity mechanism is non-biological. We offer a model in which a
competition mechanism counteracts the drive of axons along the ephrin
gradient. Will we need the axon-axon interaction mechanism based on
non-same-Eph repulsion?

@@

\section{Results}

@@

The first question we tried to answer was whether chemoaffinity alone,
implemented as gradient following, is sufficient to pattern the tectum. We set
$m_1=0.002$ and $m_2=m_3=m_4=0$ and ran the model for 300
steps.

Results are shown in Fig.\,\ref{f:ch}. Fig.\,\ref{f:ch}A indicates the colour
code used for retinal cell position; dorsal retinal cells are indicated by
green, nasal cells by red. Lightness gives position along the relevant axis so
that cells at the dorsal-nasal corner are yellow and those at the
ventral-transverse corner are black. Fig.\,\ref{f:ch}E shows the expected
layout of retinal ganglion cell axons on the tectum. The axons retain the
colour of their originating cell on the retina, as shown in
Fig.\,\ref{f:ch}A. Nasal cell axons should be arranged caudally; dorsal axons
laterally; axons should be evenly spaced and the pattern should span the
entire tectum.  Fig.\,\ref{f:ch}B--D show `fish net' plots of the axon
centroids at three timepoints. An axon's centroid is the average location of
its 4 branches. In these plots, lines are drawn between axons that are
expected to arrange themselves adjacently to one another by the end of the
simulation. The number of line crossings in this plot gives an indication of
the disorder in the arrangement. Fig.\,\ref{f:ch}G shows the positions of the
individual branches (i.e.~growth cones) at $t=300$ that contribute to the the
plot in Fig.\,\ref{f:ch}D. By this time, the topological order matches that
shown in Fig.\,\ref{f:ch}E. Opposing gradients have sorted the axons along the
two axes to form a grid of cells. However, the arrangement does not span the
full tectum. Fig.\,\ref{f:ch}F shows two pattern metrics. The first metric,
$\epsilon$, is the sum of squared distance errors of the axon centroids, with
$\mathbf{x}_{k}$, the location of the centroid of the branches of axon $k$ and
$\mathbf{x}'_{k}$ the target tectal location for axon $k$:
%
\begin{equation}\label{e:eps}
\epsilon = \sum_k |\mathbf{x}_{k} - \mathbf{x}'_{k}|^2.
\end{equation}
%
$\epsilon$ tends to 0 as the pattern approaches the target layout. However,
two patterns may have the same non-zero value for $\epsilon$, despite one
being qualitatively more disordered than the other. For this reason, we also
use the number of crossings in the fish net plot of Fig.\,\ref{f:ch}D as a
second metric, called $\eta$. The expected number of crossings in the wildtype
pattern is 0, as in Fig.\,\ref{f:ch}E, although note that in some patterns,
the expected number of crossings is non-zero. Fig.\,\ref{f:ch}H shows selected
axons (one from the centre and one from each corner of the retina) and their
position history.

@@

\section{ZEROTH DRAFT/NOTES}

>>>>

% Ability to use colour in text
\usepackage[usenames]{color}
\definecolor{grey}{rgb}{0.6, 0.6, 0.63}
\definecolor{black}{rgb}{0, 0, 0}

@@

\color{grey}
\textbf{New introduction:} Simpson and Goodhill presented a model in which the
chemoaffinity mechanism is non-biological. We offer a model in which a
competition mechanism counteracts the drive of axons along the ephrin
gradient. Will we need the axon-axon interaction mechanism based on
non-same-Eph repulsion?

@@

\color{black}
\section{Results}

@@

The first question we tried to answer was whether a gradient following
mechansim, acting alone, is sufficient to pattern the tectum. To do so, we
simulated the retina and tectum as unit square regions and defined retinal
receptor expression and tectal ligand expression across these squares
(Fig.\,\ref{f:ex}).

Each retinal ganglion cell projects growth cones (referred to
as \emph{branches}) which carry a set of 4 receptors, $r_i$ indexed by $i$ and
expressed at levels determined by the cell soma's location on the
retina. (Four ligands, $l_i$, are also expressed by RGCs, but these are not
used by the chemoaffinity simulation.) The expression levels of the receptors
vary with respect to the soma's position on the retinal surface. Each receptor
varies only with respect to one dimension.  Receptor expression gradients are
arranged in orthogonal pairs with the gradient of $r_0$ being orthogonal to
that of $r_1$. $r_2$, whose gradient is opposite to $r_0$, is orthogonal to
$r_3$.

Because there is convincing evidence that EphA and EphB receptors are
expressed in exponentially increasing
patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004},
we use an exponential form for retinal receptor expressions, in common with
other modelling
studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
We adopt the same precise form for the retinal receptor expression
as \citet{simpson_simple_2011} which is
\begin{equation} \label{e:retrcpt}
r_i(x,y) = 1.05 + 0.26 \exp(2.3 x)
\end{equation}
for even values of $i$. The opposing expressions (for even values of $i$)
are given by substituting $(1-x)$ for $x$.

The tectum expresses ligands, $L_i$, for the retinal receptors, also
in orthogonal pairs of gradients. Although several studies model tectal ligand
expression with exponential functions \citep{koulakov_stochastic_2004}, the
experimental evidence for ligand expression is more ambiguous than that for
retinal receptor expression (Supp.~Fig
\textbf{XXXXX}). We kept in mind the possibility that tectal ligand
expression may be modelled by exponential, linear or logarithmic functions,
but initially, we set it to the function
\begin{equation} \label{e:tecligexp}
L_i^{\text{exp}}(x,y) = 1.05 + 0.26 \exp(2.3 x)
\end{equation}
for odd values of $i$ and using the same substitution of $(1-x)$ for $x$ to
obtain the opposing ligand expressions for $i$ even.

While we do not explicitly name $r_0$, $r_1$,
etc.~as \emph{EphA}, \emph{EphB}, the suggestion is that $r$ includes EphA,
EphB, Ryk \citep{schmitt_wntryk_2006} and
Neogenin \citep{rajagopalan_neogenin_2004} receptors and that $L$ includes the
ephrin-A, ephrin-B, Wnt3 \citep{schmitt_wntryk_2006} and
RGM \citep{monnier_rgm_2002} ligands, each of which has been shown to play a
r\^ole in retinotectal map formation.

Figure~\ref{f:ex} shows colour maps of receptor and ligand
expression. Coordinates within the simulated retina and tectum are normalised,
with 1 corresponding to 2 mm on the tectum and on the retina for
mouse \citep{reber_relative_2004}.

\color{grey}
This means that growth cones of approximate
radius 5 $\mu$m = 0.005 mm \citep{goodhill_can_2016} have a radius $r$ =
0.0025 in the arbitrary units of the simulation. For a realistic simulation
based on mouse, the approximate RGC density in the retina is around 44000
cells per retina \citep{jeon_major_1998}. This sets the side length of the
retina to $\sqrt{44000} \approx 200$. This results in a simulation of
prodigious computational size, which although tractable (10 s per step,
requiring several hours, rather than days to run) is not convenient. To
maintain the relationship between growth cone size and cell density, we permit
a smaller number of growth cones ($400\times8=3200$) but keep the fractional
area of the branches at 0.86 ($44000 \times \pi r^2$) (if there is one cone
per axon) or 6.9 (if there are 8 cones per axon).

\color{black}
Each discrete element in Fig.\,\ref{f:ex}B projects one RGC axon with 4
branches to the tectum, carrying the receptor expression at the retinal
element's location.
At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, at time
$t$ of each branch $b$ is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where $\mathbf{M}_{b}$, the movement vector, is the sum of a chemoaffinity
effect and a `border effect', $\mathbf{B}_b$, which acts to retain branches
within the tectal region:
%
\begin{equation}
\mathbf{M}_{b} = m_1 \mathbf{G}_b + 0.5 \mathbf{B}_b
\end{equation}
%
evaluated at time $t$.

% G - chemoaffinity description
The chemoaffinity component, $\mathbf{G}_b$, is governed by receptor-ligand
signalling. The signal transmitted when a tectal ligand binds to a RGC
receptor on a branch (forward signalling) can lead to one of two effects; the
signal may induce an attraction towards the ligand-expressing region or a
repulsion away from it.
%
For axon-tectum interactions, we implemented attraction as a climbing of the
tectal ligand expression gradient and repulsion as gradient descent.
%
To determine the strength of the effect we assumed a purely linear receptor
binding model, and set the chemotactic movement vector of the branch $b$ at
location $\mathbf{x}_b$ on the tectum to be
%
\begin{equation}
\mathbf{G}_b = \sum_i^N F_i\,r_{i,b} \nabla L_i(\mathbf{x}_b)
\end{equation}
%
where $r_{i,b}$ is the receptor expression on branch $b$ for ligand-receptor
pair $i$, $\nabla L_i$ is the gradient of expression of ligand $i$ on the
tectum and $F_i$ denotes the direction of the interaction induced when a
molecule of ligand $i$ binds to a receptor $i$ molecule. $F_i$ takes the value
-1 for a repulsive interaction or 1 for an attractive interaction.
%
We assumed that all
receptor-ligand signalled interactions are repulsive ($F_i=-1, \forall i$), as
for EphA-ephrinA
coupling \citep{drescher_vitro_1995,nakamoto_topographically_1996}.
%
$m_1$ is a
scalar parameter which controls how much movement is generated for a given
receptor-ligand signal size.
% end chemoaffinity description

% border effect description
The border effect is also based on gradient following by assuming that
there is some other molecular signal which acts on all branches near the
boundary of the tectal tissue. For a branch with position $(x,y)$, $\mathbf{B}_b$ is
given by:
%
\begin{equation}
B_{b,x} = \begin{cases}
        r-x      & x<r \\
        1-r-x    & x>1-r
\end{cases}
B_{b,y} = \begin{cases}
        r-y      & y<r \\
        1-r-y    & y>1-r
\end{cases}
\end{equation}
%
Thus $\mathbf{B}_b$ is equivalent to the action of a repulsive signalling
molecule expressed around the border of the tectum, whose expression increases
quadratically outside the tectum and affects any branch touching (or outside)
the boundary.
% end border effect description

% Initial conditions
Branches were initially randomly distributed in a stripe at the rostral side
of the tectum at the start of each simulation. Each RGC axon was assigned a
random initial position coordinate:
\begin{equation}
\mathbf{x}_{\text{axon},t=0} = (U(0,1), U(-0.2,0))
\end{equation}
where U(p,q) is a number selected from a random uniform distribution in the
range $[p,q)$. Each of the 4 branches per RGC axon was given its parent
axon's initial position, plus a randomly generated offset with coordinates
derived from a normal distribution of mean 0 and standard deviation 0.1.
% end initial conditions

The results of running this simulation for 300 time steps are shown in
Fig.\,\ref{f:ch}. Fig.\,\ref{f:ch}A indicates the colour code used for retinal
cell position; dorsal retinal cells are indicated by green, nasal cells by
red. Lightness gives position along the relevant axis so that cells at the
dorsal-nasal corner are yellow and those at the ventral-transverse corner are
black. Fig.\,\ref{f:ch}E shows the expected layout of retinal ganglion cell
axons on the tectum. The axons retain the colour of their originating cell on
the retina, as shown in Fig.\,\ref{f:ch}A. Nasal cell axons should be arranged
caudally; dorsal axons laterally; axons should be evenly spaced and the
pattern should span the entire tectum.  Fig.\,\ref{f:ch}B--D show `fish net'
plots of the axon centroids at three timepoints. An axon's centroid is the
average location of its 4 branches. In these plots, lines are drawn between
axons that are expected to arrange themselves adjacently to one another by the
end of the simulation. The number of line crossings in this plot gives an
indication of the disorder in the arrangement. Fig.\,\ref{f:ch}G shows the
positions of the individual branches (i.e.~growth cones) at $t=300$ that
contribute to the the plot in Fig.\,\ref{f:ch}D. By this time, the topological
order matches that shown in Fig.\,\ref{f:ch}E. Opposing gradients have sorted
the axons along the two axes to form a grid of cells. However, the arrangement
does not span the full tectum. Fig.\,\ref{f:ch}F shows two pattern
metrics. The first metric, $\epsilon$, is the sum of squared distance errors
of the axon centroids, with $\mathbf{x}_{k}$, the location of the centroid of
the branches of axon $k$ and $\mathbf{x}'_{k}$ the target tectal location for
axon $k$:
%
\begin{equation}\label{e:eps}
\epsilon = \sum_k |\mathbf{x}_{k} - \mathbf{x}'_{k}|^2.
\end{equation}
%
$\epsilon$ tends to 0 as the pattern approaches the target layout. However,
two patterns may have the same non-zero value for $\epsilon$, despite one
being qualitatively more disordered than the other. For this reason, we also
use the number of crossings in the fish net plot of Fig.\,\ref{f:ch}D as a
second metric, called $\eta$. The expected number of crossings in the wildtype
pattern is 0, as in Fig.\,\ref{f:ch}E, although note that in some patterns,
the expected number of crossings is non-zero. Fig.\,\ref{f:ch}H shows selected
axons (one from the centre and one from each corner of the retina) and their
position history.

@@

\section{Discussion}
Discussion goes here.

\color{grey}
\section{ZEROTH DRAFT/NOTES}
